RNA-seq from 4 cell populations

### 2021-07-28 deconvolution/fans4/molec_profiling.Rmd
Initial commit
freeze 2

@@

```{r load.packages, echo=FALSE, message=FALSE, results='hide', cache=FALSE}
suppressPackageStartupMessages({
library(variancePartition)
library(limma)
library(edgeR)
library(mvIC)
library(reshape2)
library(qvalue)
library(data.table)
library(BiocParallel)
library(knitr)
library(org.Hs.eg.db)
library(kableExtra)
library(ggplot2)
library(cowplot)
library(scran)
library(SummarizedExperiment)
library(gridExtra)
library(openxlsx)
library(synapser)
synLogin()
library(EnsDb.Hsapiens.v86)
detach("package:EnsDb.Hsapiens.v86", unload=TRUE)
library(EnsDb.Hsapiens.v86)
ensdb = EnsDb.Hsapiens.v86
})

@@

### Reading RNA-seq data
```{r read.data, message=TRUE}
Ids = c('syn22321012', 'syn22321013', 'syn22321014', 'syn22321016')
# file = "/sc/arion/projects/CommonMind/roussp01a/MOLECULAR_PROFILING/rnaseq/step1-merged/featureCount/featureCount.gene.tsv.gz"
file = synGet('syn22321012')$path
df_featureCounts_gene = fread( file )

@@

### RNA-seq QC metrics
```{r QC, fig.width=18, fig.height=74}
# file = '/sc/arion/projects/CommonMind/roussp01a/MOLECULAR_PROFILING/rnaseq/step1-merged/QCmetrics/qc_metrics.RDS'
file = synGet('syn22321016')$path
qc_metrics.in = t(readRDS( file ))
colnames(qc_metrics.in) = qc_metrics.in[1,]
qc_metrics.in = as.data.frame(qc_metrics.in[-c(1:2),], stringsAsFactors=FALSE)
qc_metrics = apply(qc_metrics.in, 2, as.numeric)
rownames(qc_metrics) = gsub('-', '.', rownames(qc_metrics.in))

@@

# add information about each sample
info = merge(info, metadata1, by.x="DissectionID", by.y='Dissection.ID')
info = merge(info, qc_metrics, by.x="Name", by.y='row.names')
rownames(info) = info$Name

@@

# set order to be the same as geneCounts
info = info[match(colnames(geneCounts), info$Name),]

@@

df = reshape2::melt(geneExpr)
df = merge(df, info[,'CellType',drop=FALSE], by.x = "Var2", by.y="row.names")

@@

### PCA
```{r PCA}

dcmp = prcomp( t(geneExpr), scale=TRUE)

@@

xlab = paste0('PC1 (', format(varFrac[1]*100, digits=3), '%)')
ylab = paste0('PC2 (', format(varFrac[2]*100, digits=3), '%)')
ggplot(df_pca, aes(PC1, PC2, color=CellType)) + geom_point() + theme_bw(15) + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + scale_color_manual("Cell Type", values=colorCodes) + xlab(xlab) + ylab(ylab)

xlab = paste0('PC2 (', format(varFrac[2]*100, digits=3), '%)')
ylab = paste0('PC3 (', format(varFrac[3]*100, digits=3), '%)')
ggplot(df_pca, aes(PC2, PC3, color=CellType)) + geom_point() + theme_bw(15) + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + scale_color_manual("Cell Type", values=colorCodes) + xlab(xlab) + ylab(ylab)
```

@@

### variancePartition
```{r vp}
form = ~ (1|CellType) + (1|Dx) + (1|CellType:Dx) + (1|Individual.ID) + (1|Gender)
vp = fitExtractVarPartModel( geneExpr, form, info, BPPARAM=SnowParam(6, progressbar=TRUE))

@@

## Differential expression analysis
### Cell type
```{r DE_process}
form = ~ 0 + CellType + (1|Individual.ID) + (1|Gender)

@@

# Differential expression
file = Roussos_MSSM_MolProf_FANS_4_diffExpression.RDS

>>>>

cd /Users/gabrielhoffman/workspace/repos/CMC/covarr-de/deconvolution/fans4
R

rmarkdown::render("molec_profiling.Rmd", output_dir='./', intermediates_dir='./')

@@

```{r load.packages, echo=FALSE, message=FALSE, results='hide', cache=FALSE}
suppressPackageStartupMessages({
library(variancePartition)
library(limma)
library(edgeR)
library(mvIC)
library(reshape2)
library(qvalue)
library(data.table)
library(BiocParallel)
library(knitr)
library(org.Hs.eg.db)
library(kableExtra)
library(ggplot2)
library(cowplot)
library(scran)
library(SummarizedExperiment)
library(gridExtra)
library(openxlsx)
library(synapser)
library(githubr)
synLogin()
library(EnsDb.Hsapiens.v86)
detach("package:EnsDb.Hsapiens.v86", unload=TRUE)
library(EnsDb.Hsapiens.v86)
ensdb = EnsDb.Hsapiens.v86
})

@@

### Reading RNA-seq data
```{r read.data, message=TRUE}
Ids = c('syn22321012', 'syn22321013', 'syn22321014', 'syn22321016')

# Read in RNA-seq featureCounts
###############################

# Phase 1
file = synGet('syn22321012')$path
df_featureCounts_gene = fread( file )

# Phase 2+3
# syn26000845 is really phase2 again
ids = c('syn26000845','syn25976117', 'syn25976127')[-1]
df_read = lapply(ids, function(id){
  df = fread(synGet(id)$path)
  colnames(df) = gsub("^Sample_", '', colnames(df))
  colnames(df) = gsub("-", '.', colnames(df))
  df
  })

# save batch info
df_batch = data.frame(Name = colnames(df_featureCounts_gene)[-c(1:6)], Batch=1)
df_batch = rbind(df_batch, data.frame(Name = colnames(df_read[[1]][,-c(1:6)]), Batch=2))
df_batch = rbind(df_batch, data.frame(Name = colnames(df_read[[2]][,-c(1:6)]), Batch=3))
df_batch$Batch = as.character(df_batch$Batch)

# sapply(df_read, ncol)
df_featureCounts_gene = cbind(df_featureCounts_gene, 
                          df_read[[1]][,-c(1:6)], 
                          df_read[[2]][,-c(1:6)])

@@

### RNA-seq QC metrics
```{r QC, fig.width=18, fig.height=74}
# Phase 1 
# file = '/sc/arion/projects/CommonMind/roussp01a/MOLECULAR_PROFILING/rnaseq/step1-merged/QCmetrics/qc_metrics.RDS'
file = synGet('syn22321016')$path
qc_metrics.in = t(readRDS( file ))
colnames(qc_metrics.in) = qc_metrics.in[1,]
qc_metrics.in = as.data.frame(qc_metrics.in[-c(1:2),], stringsAsFactors=FALSE)
qc_metrics = apply(qc_metrics.in, 2, as.numeric)
rownames(qc_metrics) = gsub('-', '.', rownames(qc_metrics.in))

# Phase 2+3
ids = c('syn26000850', 'syn25976122', 'syn25976136')[-1]
df_read = lapply(ids, function(id){
  qc.in = read.table(synGet(id)$path)
  rownames(qc.in) = gsub('-', '.', rownames(qc.in))
  qc.in
  })

# # sapply(df_read, ncol)
# df_featureCounts_gene = cbind(df_featureCounts_gene, 
#                           df_read[[1]][,-c(1:6)], 
#                           df_read[[2]][,-c(1:6)])

@@

# add information about each sample
info = merge(info, metadata1, by.x="DissectionID", by.y='Dissection.ID', all.x=TRUE)
info = merge(info, qc_metrics, by.x="Name", by.y='row.names', all.x=TRUE)
rownames(info) = info$Name

@@

# Merge with batch info
info = merge(info, df_batch, by="Name")
rownames(info) = info$Name

# set order to be the same as geneCounts
# Must be last change to info
info = info[match(colnames(geneCounts), info$Name),]

@@

df = data.table(reshape2::melt(geneExpr))
setkey(df, "Var2")
df = merge(df, info[,c("Name", 'CellType')], by.x = "Var2", by.y="Name")

@@

### PCA
```{r PCA, fig.width=10, fig.height=5}
dcmp = prcomp( t(geneExpr), scale=TRUE)

@@

xlab = paste0('PC1 (', format(varFrac[1]*100, digits=3), '%)')
ylab = paste0('PC2 (', format(varFrac[2]*100, digits=3), '%)')
fig1 = ggplot(df_pca, aes(PC1, PC2, color=CellType, shape=Batch)) + geom_point() + theme_bw(15) + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + scale_color_manual("Cell Type", values=colorCodes) + xlab(xlab) + ylab(ylab)

xlab = paste0('PC2 (', format(varFrac[2]*100, digits=3), '%)')
ylab = paste0('PC3 (', format(varFrac[3]*100, digits=3), '%)')
fig2 = ggplot(df_pca, aes(PC2, PC3, color=CellType, shape=Batch)) + geom_point() + theme_bw(15) + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + scale_color_manual("Cell Type", values=colorCodes) + xlab(xlab) + ylab(ylab)

plot_grid(fig1, fig2 + theme(legend.position="none"), axis="tbr", align="v")
```

#### Batch corrected
```{r PCA.batch, fig.width=10, fig.height=5}
fit = dream(geneExpr, ~ Batch, info)

dcmp = prcomp( t(residuals(fit)), scale=TRUE)

df_pca = merge(df_pca, info[,c("Name", "CellType", "Batch")], by.x="row.names", by.y="Name")

xlab = paste0('PC1 (', format(varFrac[1]*100, digits=3), '%)')
ylab = paste0('PC2 (', format(varFrac[2]*100, digits=3), '%)')
fig1 = ggplot(df_pca, aes(PC1, PC2, color=CellType, shape=Batch)) + geom_point() + theme_bw(15) + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + scale_color_manual("Cell Type", values=colorCodes) + xlab(xlab) + ylab(ylab)

xlab = paste0('PC2 (', format(varFrac[2]*100, digits=3), '%)')
ylab = paste0('PC3 (', format(varFrac[3]*100, digits=3), '%)')
fig2 = ggplot(df_pca, aes(PC2, PC3, color=CellType, shape=Batch)) + geom_point() + theme_bw(15) + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + scale_color_manual("Cell Type", values=colorCodes) + xlab(xlab) + ylab(ylab)

plot_grid(fig1, fig2 + theme(legend.position="none"), axis="tbr", align="v")
```

```{r predict.celltype}
library(nnet)
library(PRROC)

hush=function(code){
  sink("NUL") # use /dev/null in UNIX
  tmp = code
  sink()
  return(tmp)
}

df_class = lapply(1:10, function(k){

  fit = hush(multinom(df_pca$CellType ~ X[,1:k,drop=FALSE]))

  data.table(k =k, n_mismatch = sum(predict(fit) != df_pca$CellType))
  })
df_class = do.call(rbind, df_class)

fig1 = ggplot(df_class, aes(factor(k), n_mismatch)) + geom_bar(stat="identity") + theme_classic() + scale_y_continuous(expand=c(0,0)) + ggtitle("Ability to predict each cell type") + theme(aspect.ratio=1)

df_class = lapply(1:10, function(k){

  fit = hush(multinom(df_pca$CellType ~ X[,1:k,drop=FALSE]))

  Y_pred = predict(fit, type="probs")

  df_class = lapply( levels(df_pca$CellType), function(CT){

    y0 = Y_pred[df_pca$CellType == CT,CT]
    y1 = Y_pred[df_pca$CellType != CT,CT]

    pr = pr.curve(scores.class0 = y0, scores.class1 = y1, curve=TRUE, rand.compute=TRUE)

    data.table(k = k, CellType = CT, AUPR = pr$auc.integral, AUPR.rand = pr$rand$auc.integral)
    })
  do.call(rbind, df_class)
  })
df_class = do.call(rbind, df_class)

fig2 = ggplot(df_class, aes(k, AUPR, color=CellType)) + geom_point() + geom_line() + theme_classic() + scale_color_manual(values = colorCodes) + scale_x_continuous(breaks = 1:10) + ggtitle("Ability to predict each cell type") + theme(aspect.ratio=1)

plot_grid(fig1, fig2, axis="tbr", align="v")
```

@@

### variancePartition
```{r vp}
form = ~ (1|CellType) + (1|Dx) + (1|CellType:Dx) + (1|Individual.ID) + (1|Gender) + (1|Batch)
vp = fitExtractVarPartModel( geneExpr, form, info, BPPARAM=SnowParam(6, progressbar=TRUE))

@@

## Differential expression analysis
### Cell type
```{r DE_process}
form = ~ 0 + CellType + (1|Individual.ID) + (1|Gender) + (1|Batch)

@@

# Differential expression
file = "Roussos_MSSM_MolProf_FANS_4_diffExpression.RDS"
